\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2010-03-04 papers/jss/dpx4Rev.Rnw
Minor changes in paper and getdf for transInit models (not done yet ...)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@382 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

	\pkg{depmixS4} implements a general framework for defining and
	estimating dependent mixture models in the \proglang{R}
	programming language \citep{R2009}.  This includes standard Markov
	models, latent/hidden Markov models, and latent class and finite
	mixture distribution models.  The models can be fitted on mixed
	multivariate data with distributions from the \code{glm} family,
	the logistic multinomial, or the multivariate normal distribution.
	Other distributions can be added easily, and an example is
	provided with the exgaus distribution.  Parameters are estimated by
	the EM algorithm or, when (linear) constraints are imposed on the
	parameters, by direct numerical optimization with the
	\pkg{Rdonlp2} routine.}

\Keywords{hidden Markov model, dependent mixture model, mixture model}

\Plainkeywords{hidden Markov model, dependent mixture model, mixture model} %% without formatting
%% at least one keyword must be supplied

@@

The data considered here have the general form $\vc{O}_{1:T}=
(O_{1}^{1}, \ldots, O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$,
\ldots, $O_{T}^{1}, \ldots, O_{T}^{m})$ for an $m$-variate time series
of length $T$.  In the following, we use $\vc{O}_{t}$ as shorthand for
$O_{t}^{1}, \ldots, O_{t}^{m}$.  As an example, consider a time series
of responses generated by a single participant in a psychological
response time experiment.  The data consists of three variables,
response time, response accuracy, and a covariate which is a pay-off
variable reflecting the relative reward for speeded and/or accurate
responding.  These variables are measured on 168, 134 and 137
occasions respectively (the first part of this series is plotted in
Figure~\ref{fig:speed}).  These data are more fully described in
\citet{Dutilh2010}.

\begin{figure}[htbp]
\begin{center}
<<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=
library(depmixS4)
data(speed)
plot(as.ts(speed[1:168,]),main="Speed-accuracy trade-off")
@
	 \caption{Response times (rt), accuracy (corr) and pay-off values (Pacc) for
	  the first series of responses in dataset \code{speed}.}
	  \label{fig:speed}

@@

The latent Markov model is usually associated with data of this type,
in particular for multinomially distributed responses.  However,
commonly employed estimation procedures \citep[e.g.,][]{Pol1996}, are
not suitable for long time series due to underflow problems.  In
contrast, the hidden Markov model is typically only used for `long'
univariate time series \citep[][chapter~1]{Cappe2005}.  We use the
term ``dependent mixture model'' because one of the authors (Ingmar
Visser) thought it was time for a new name to relate these
models\footnote{Only later did I find out that \citet{Leroux1992}
already coined the term dependent mixture models in an application
with hidden Markov mixtures of Poisson count data.}.

@@

%In the next paragraphs, the likelihood
%equation and estimation procedures for the dependent mixture model are
%described for data of the above form.

%The likelihood of the dependent mixture model conditional on the
%unknown (or hidden) state sequence $\vc{S}_{1:T} = (S_1,\ldots,S_T)$ 
%and the model is given
%by:
%\begin{equation}
%	\Prob( \vc{O}_{1:T} | \vc{S}_{1:T}, \greekv{\theta} ) = 
%	\prod_{t=1}^{T}  \Prob( \vc{O}_{t} | S_{t},  \greekv{\theta}),
%\end{equation}
%where $\greekv{\theta}$ is the parameter vector of the model.  To arrive
%at the marginal likelihood of the data for parameter vector
%$\greekv{\theta}$, i.e.\ unconditional on the state sequence, we need to sum
%above likelihood over all possible state sequences. This likelihood 
%is written as: 
%\begin{equation}
%	\Prob(\vc{O}_{1:T}|\greekv{\theta}) =  
%	\sum_{all Ss} \pi_{i}(\vc{z}_1) \vc{b}_{1}(\vc{O}_{1}) 
%	\prod_{t=1}^{T-1} a_{ij}(\vc{z}_{t}) \vc{b}_{j}(\vc{O}_{t+1}|\vc{z}_{t+1}),
%\end{equation}
%where we have the following elements:
%\begin{enumerate}
%	
%	\item $S_{t}$ is an element of $\mathcal{S}=\{1\ldots n\}$, a set
%	of $n$ latent classes or states; we write for short $S_{i}$ to
%	denote $S_{t}=i$.
%	
%	\item $\pi_{i}(\vc{z}_1) = \Prob(S_1 = i|\vc{z}_1)$,
%	giving the probability of class/state $i$ at time $t=1$ with
%	covariate $\vc{z}_1$.
%	
%	\item $a_{ij}(\vc{z}_t) = \Prob(S_{t+1}=j|S_{t}=i,\vc{z}_t)$,
%	provides the probability of a transition from state $i$ to state
%	$j$ with covariate $\vc{z}_t$,
%	
%	\item $\vc{b}_{j}$ is a vector of observation densities
%	$b_j^k(\vc{z}_t) = \Prob(O_{t}^k|S_t = j, \vc{z}_t)$ that provide the
%	conditional densities of observations $O_{t}^k$ associated with
%	latent class/state $j$ and covariate $\vc{z}_t$, $j=1, \ldots, n$,
%	$k=1, \ldots, m$. 
%\end{enumerate}

% The dependent mixture model is defined by the following equations: 
% %\begin{align}
% %	S_{t} &= A S_{t-1}, t=2, \ldots, T \\
% %	O_{t} &= b(O_{t}|S_{t}), 
% %\end{align}
% \begin{align}
% 	Pr(S_{t} = j) &= \sum_{i=1}^n a_{ij}(\vc{x}_t) Pr(S_{t-1} = i), t=2, \ldots, T \\
% 	p(O_{t}^k|S_t &= j, \vc{x}_t) = b_j^k(\vc{x}_t)
% \end{align}
%where $S_{t}$ is a sequence of hidden states, $A$ is a transition
%matrix, $O_{t}$ is an (possibly multivariate) observation and $b$ is a
%density function for $O_{t}$ conditional on the hidden state $S_{t}$.

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM
algorithm, parameters are estimated by iteratively maximising the
expected joint log-likelihood of the parameters given the observations and
states.  Let $\greekv{\theta} = (\greekv{\theta}_1,
\greekv{\theta}_2,\greekv{\theta}_3)$ be the general parameter vector
consisting of three subvectors with parameters for the prior model,
transition model, and response model respectively.  The joint
log-likelihood can be written as
\begin{equation}
\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(\vc{O}_t|S_t,\vc{z}_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
Expectation step, we replace these with their expected values given a set of 
(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $\vc{O}_{1:T}$. 
The expected log-likelihood:
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta}))
\end{equation}
can be written as
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
\log \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
\end{multline}
where the expected values $\xi_t(j,k) = P(S_t = k, S_{t-1} =
j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) =
P(S_t = j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ can be
computed effectively by the forward-backward algorithm \citep[see
e.g.,][]{Rabiner1989}.  The Maximisation step consists of the
maximisation of (\ref{eq:Q}) for $\greekv{\theta}$.  As the right hand
side of (\ref{eq:Q}) consists of three separate parts, we can maximise
separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and
$\greekv{\theta}_3$.  In common models, maximisation for
$\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the
\code{nnet.default} routine in the \pkg{nnet} package
\citep{Venables2002}, and maximisation for $\greekv{\theta}_3$ by the
standard \code{glm} routine.  Note that for the latter maximisation,
the expected values $\gamma_t(j)$ are used as prior weights of the
observations $O^k_t$.

%The maximising values of $a_{jk}$ are \cite<e.g.,>{Rabiner89}
%\begin{equation}
%\label{eq:MLA}
%\hat{a}_{jk} = \frac{1}{N(T-1)} \sum_{i=1}^N \sum_{t=2}^T  \frac{\xi^i_t(j,k)}{\gamma_{t-1}^i(j)}.
%\end{equation}

%Fixing certain elements of $\mat{A}$ to 0, as in the DT version, does
%not affect the estimation of the other elements.  When elements
%$a_{jk}$ are assumed identical, we simply extend the summation in
%(\ref{eq:MLA}) to include all those elements, changing the
%denominator $N(T-1)$ accordingly.

%To estimate $\greekv{\lambda}$, we note that the term containing this
%parameter has the form of a weighted likelihood, where
%$\gamma_t^i(j)$ can be interpreted as the number of replications of
%$r^i_t$.  Hence, we can rely on standard ML estimates of the logistic
%regression coefficients $\lambda_i$, using the values $\gamma_t^i(j)$
%as ``case weights'' \cite<e.g.,>{Agresti02,McCullagh83}\footnote{To
%be more specific, we replicate each observation $r^i_t$ a total of 6
%times, once for each of the states besides the random state (which
%offers no information regarding $\lambda_i$).  For the $j$-th
%replication (corresponding to the $j$-th state), we used
%$v_j(\vec{x}_t)$ as a predictor variable and $\gamma^i_t(j)$ as a
%case weight.  All these replications were used to obtain the maximum
%likelihood estimate of $\lambda$ from a single GLM, using the
%``glm.fit'' function in R \cite{R}.}.  The resulting parameter
%estimates are then used in a new Expectation step, followed by
%another Maximisation step, until convergence.  Under mild regularity
%conditions, it can be shown that the EM algorithm converges to a
%(local) maximum of the likelihood \cite{EM}.

The EM algorithm however has some drawbacks.  First, it can be slow to
converge towards the end of optimization.  Second, applying
constraints to parameters can be problematic; in particular, EM can
lead to wrong parameter estimates when applying constraints.  Hence,
in \pkg{depmixS4}, EM is used by default in unconstrained models, but
otherwise, direct optimization is used.  Two options are available for
direct optimization using package \pkg{Rdonlp2}
\citep{Tamura2009,Spellucci2002}, or package \pkg{Rsolnp}.  Both
packages can handle general linear (in)equality constraints, and
optionally also non-linear constraints.

%Need some more on EM and how/why it is justified to do separate weighted
%fits of the response models and transition and prior models.

%Also mention use of glm, nnet and possibly other packages that we use.

>>>>

	\pkg{depmixS4} implements a general framework for defining and
	estimating dependent mixture models in the \proglang{R}
	programming language \citep{R2009}.  This includes standard Markov
	models, latent/hidden Markov models, and latent class and finite
	mixture distribution models.  The models can be fitted on mixed
	multivariate data with distributions from the \code{glm} family,
	the logistic multinomial, or the multivariate normal distribution.
	Other distributions can be added easily, and an example is
	provided with the exgaus distribution.  Parameters are estimated by
	the EM algorithm or, when (linear) constraints are imposed on the
	parameters, by direct numerical optimization with the
	\pkg{Rdonlp2} or \pkg{Rsolnp} routines.}

\Keywords{hidden Markov model, dependent mixture model, mixture 
model, constraints}

\Plainkeywords{hidden Markov model, dependent mixture model, mixture 
model, constraints} %% without formatting
%% at least one keyword must be supplied

@@

The data considered here have the general form $\vc{O}_{1:T}=
(O_{1}^{1}, \ldots, O_{1}^{m}$, $O_{2}^{1}, \ldots, O_{2}^{m}$,
\ldots, $O_{T}^{1}, \ldots, O_{T}^{m})$ for an $m$-variate time series
of length $T$.  In the following, we use $\vc{O}_{t}$ as shorthand for
$O_{t}^{1}, \ldots, O_{t}^{m}$.  As an example, consider a time series
of responses generated by a single participant in a psychological
response time experiment.  The data consists of three variables:
response time, response accuracy, and a covariate which is a pay-off
variable reflecting the relative reward for speeded and/or accurate
responding.  These variables are measured on 168, 134 and 137
occasions respectively (the first series of 168 trials is plotted in
Figure~\ref{fig:speed}).  These data are more fully described in
\citet{Dutilh2010}, and in the next section a number of example models
for these data is described.

\begin{figure}[htbp]
\begin{center}

	<<fig=TRUE,echo=FALSE,height=5,width=7,eps=FALSE>>=
	library(depmixS4)
	data(speed)
	plot(as.ts(speed[1:168,]),main="Speed-accuracy trade-off")
	@

	\caption{Response times (rt), accuracy (corr) and pay-off 
	values (Pacc) for the first series of responses in dataset 
	\code{speed}.}

	\label{fig:speed}

@@

The latent Markov model is usually associated with data of this type,
in particular for multinomially distributed responses.  However,
commonly employed estimation procedures \citep[e.g.,][]{Pol1996}, are
not suitable for long time series due to underflow problems.  In
contrast, the hidden Markov model is typically only used for `long'
univariate time series \citep[][chapter~1]{Cappe2005}.  We use the
term "dependent mixture model" because one of the authors (Ingmar
Visser) thought it was time for a new name to relate these
models\footnote{Only later did I find out that \citet{Leroux1992}
already coined the term dependent mixture models in an application
with hidden Markov mixtures of Poisson count data.}.

@@

@@

Parameters are estimated in \pkg{depmixS4} using the EM algorithm or
through the use of a general Newton-Raphson optimizer.  In the EM
algorithm, parameters are estimated by iteratively maximising the
expected joint log-likelihood of the parameters given the observations and
states.  Let $\greekv{\theta} = (\greekv{\theta}_1,
\greekv{\theta}_2,\greekv{\theta}_3)$ be the general parameter vector
consisting of three subvectors with parameters for the prior model,
transition model, and response models respectively.  The joint
log-likelihood can be written as:
\begin{equation}
\log \Prob(\vc{O}_{1:T}, \vc{S}_{1:T}|\vc{z}_{1:T},\greekv{\theta}) = \log 
\Prob(S_1|\vc{z}_{1},\greekv{\theta}_1) 
+ \sum_{t=2}^{T} \log \Prob(S_t|S_{t-1},\vc{z}_{t-1},\greekv{\theta}_2) 
+ \sum_{t=1}^{T} \log \Prob(\vc{O}_t|S_t,\vc{z}_t,\greekv{\theta}_3)
\end{equation}
This likelihood depends on the unobserved states $\vc{S}_{1:T}$. In the 
Expectation step, we replace these with their expected values given a set of 
(initial) parameters $\greekv{\theta}' = (\greekv{\theta}'_1, 
\greekv{\theta}'_2,\greekv{\theta}'_3)$ and observations $\vc{O}_{1:T}$. 
The expected log-likelihood:
\begin{equation}
Q(\greekv{\theta},\greekv{\theta}') = E_{\greekv{\theta}'} 
(\log \Prob(\vc{O}_{1:T},\vc{S}_{1:T}|\vc{O}_{1:T},\vc{z}_{1:T},\greekv{\theta})),
\end{equation}
can be written as:
\begin{multline}
\label{eq:Q}
Q(\greekv{\theta},\greekv{\theta}') = 
\sum_{j=1}^n \gamma_1(j) \log \Prob(S_1=j|\vc{z}_1,\greekv{\theta}_1) \\ 
+ \sum_{t=2}^T \sum_{j=1}^n \sum_{k=1}^n \xi_t(j,k) \log \Prob(S_t = k|S_{t-1} 
= j,\vc{z}_{t-1},\greekv{\theta}_2)  \\
 + \sum_{t=1}^T \sum_{j=1}^n \sum_{k=1}^m \gamma_t(j) 
\log \Prob(O^k_t|S_t=j,\vc{z}_t,\greekv{\theta}_3),
\end{multline}
where the expected values $\xi_t(j,k) = P(S_t = k, S_{t-1} =
j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ and $\gamma_t(j) =
P(S_t = j|\vc{O}_{1:T}, \vc{z}_{1:T},\greekv{\theta}')$ can be
computed effectively by the forward-backward algorithm \citep[see
e.g.,][]{Rabiner1989}.  The Maximisation step consists of the
maximisation of (\ref{eq:Q}) for $\greekv{\theta}$.  As the right hand
side of (\ref{eq:Q}) consists of three separate parts, we can maximise
separately for $\greekv{\theta}_1$, $\greekv{\theta}_2$ and
$\greekv{\theta}_3$.  In common models, maximisation for
$\greekv{\theta}_1$ and $\greekv{\theta}_2$ is performed by the
\code{nnet.default} routine in the \pkg{nnet} package
\citep{Venables2002}, and maximisation for $\greekv{\theta}_3$ by the
standard \code{glm} routine.  Note that for the latter maximisation,
the expected values $\gamma_t(j)$ are used as prior weights of the
observations $O^k_t$.

The EM algorithm however has some drawbacks.  First, it can be slow to
converge towards the end of optimization.  Second, applying
constraints to parameters can be problematic; in particular, EM can
lead to wrong parameter estimates when applying constraints.  Hence,
in \pkg{depmixS4}, EM is used by default in unconstrained models, but
otherwise, direct optimization is used.  Two options are available for
direct optimization using package \pkg{Rdonlp2}
\citep{Tamura2009,Spellucci2002}, or package \pkg{Rsolnp}
\citep{Ghalanos2010}.  Both packages can handle general linear
(in)equality constraints (and optionally also non-linear constraints).
